---
title: "Poorboy et al. Bite force"
author: "Erika Barthelmess"
date: "`r Sys.Date()`"
output: html_document
---

Working on data and analysis for "An Introduction to R for Mammalogists" workshop with Erin and Sean at ASM/IMC 2023 in Anchorage.

Data come from [this paper](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0284094) and can be accessed [from dryad here](https://datadryad.org/stash/dataset/doi:10.5061/dryad.kd51c5bbg)

I'm to cover wrangling and tidying of data
* dplyr "verbs"
* pivot_longer() and pivot_wider()
* getting data into a data frame/tibble format

Start with some data
```{r}
#read in data
library(here)
library(readxl)
DF <- read_excel(here("Data","Poorboyetal_Hybridization_Measurements.xls"))
```
Now turn on some packages we will need
```{r}
library(dplyr)
#library(tidyverse)
```

Let's look at the basic structure of our dataset
```{r}
str(DF)
summary(DF)
```
To wrangle data with the tidyverse, we use the `dplyr` package "verbs" via these functions:
* `select()` grabs columns
* `mutate()` creates new variables
* `filter()` grabs rows matching particular conditions
* `slice()` grabs rows NOT SURE I'll do this one

* `summarise()` provides a summary of values
* `group_by()` allows you to use any of the above operations "by group"

Make sure to pipe!

https://dplyr.tidyverse.org/articles/dplyr.html


Probabaly not time for pivot_longer() or pivot_wider().

