---
title: "My First Markdown"
author: "Sean"
date: "2023-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My first RMarkdown
The RMarkdown files are great ways to generate documents that need to include results from analysis. You can do the analysis in chunks of code and then provide text and even reference outputs in the open sections. Some of the options are a bit beyond this workshop, but a bunch of helpful hints are here  (https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

Note that some of these options will not appear like this within the RMarkdown file... 

*italics*

**bold**

***italics and bold***

...those little start will disappear and the text will appear as expected.

## Let's look at our dataset...

Mammals are so cool. Let's look at bite force...

These data come from here (https://doi:10.5061/dryad.kd51c5bbg) and are associated with this manuscript (https://doi.org/10.1371/journal.pone.0284094).


```{r}
library(here)
Measurements <- readxl::read_excel(here("data", "Poorboyetal_Hybridization_Measurements.xls"))

Measurements <- readxl::read_excel("C:/Users/Sean Maher/Documents/GitHub/Intro-R-Mammalogists/Data/Poorboyetal_Hybridization_Measurements.xls")

names(Measurements)

# FYI... Words in column names can be separated by a space, but you need to wrap the name with this character -> `

## Getting help ####
# ?plot
# help(plot)

# plotting #####

#### Plotting two variables as a formula ####
plot(Incisor_BiteForce~LogMass, data=Measurements)

m1 <- lm(Incisor_BiteForce~LogMass, data=Measurements)

abline(m1)

(m1_summary <- summary(m1))

```

We can reference our objects in the text field like so...

The linear model included a slope of `r round(m1$coefficients[2], 2)` that was different than zero (*t*= `r round(m1_summary$coefficients[2,3],2)`, df=`r m1_summary$df[2]`, *P* = `r m1_summary$coefficients[2,4]`).

## Sequential versus Piping when writing code

Generally speaking the syntax for code follows flow of creating an object based on a set of instructions designed as a function. Each function has a set of arguments that can modify or detail the action.

```{r}
## Sequential
Females <- subset(Measurements, Sex=="F") #Keep track of F and T, which are short for FALSE and TRUE, respectively
Males <- subset(Measurements, Sex!="F")

Aver_Mass_F<-mean(Females$Mass, na.rm=T) #Notice the dollar sign indicating the column
Aver_Mass_M<-mean(Males$Mass, na.rm=T)

## Piped
Aver_Mass_F <- Measurements %>%
  subset(Sex=="F") %>%
  pull(Mass)%>% #Notice this extra function call here to indicate which column
  mean(na.rm=T)

Aver_Mass_M <- Measurements %>%
  subset(Sex=="M") %>%
  pull(Mass)%>%
  mean(na.rm=T)

```

