---
title: "My First Markdown"
author: "Sean"
date: "2024-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My first RMarkdown
The RMarkdown files are great ways to generate documents that need to include results from analysis. You can do the analysis in chunks of code and then provide text and even reference outputs in the open sections. Some of the options are a bit beyond this workshop, but a bunch of helpful hints are here  (https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

Note that some of these options will not appear like this within the RMarkdown file... 

*italics*

**bold**

***italics and bold***

...those little start will disappear and the text will appear as expected.

## Let's look at our dataset...

Mammals are so cool. Let's look at biteforce...

```{r}
Measurements <- read.csv("data/cinerea.csv")

names(Measurements)

## Getting help ####
# ?plot
# help(plot)

# plotting #####

#### Plotting two variables as a formula ####
plot(Mass~Min_Temp, data=Measurements)

m1 <- lm(Incisor_BiteForce~LogMass, data=Measurements)

abline(m1)

(m1_summary <- summary(m1))

```

We can reference our objects in the text field like so...

The linear model included a slope of `r round(m1$coefficients[2], 2)` that was different than zero (*t*= `r round(m1_summary$coefficients[2,3],2)`, df=`r m1_summary$df[2]`, *P* = `r m1_summary$coefficients[2,4]`).

